A family of accessibility measures derived from spatial interaction principles

### 2025-08-28 paper_plos/paper_plos.qmd
rafa's comments and edits

Firstly, there is a group of articles within this stream that cite Wilson [@wilson1971] exclusively as attribution for using context-dependent travel cost functions. This trend is common: for instance it is done in [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining]. However, these works do not engage with spatial interaction beyond this attribution.

Secondly, a subset of literature explicitly associates spatial interaction, as defined in Wilson [@wilson1971], with accessibility’s potential for spatial interaction--but only superficially. These works acknowledge the conceptual link, but do not go beyond this recognition in the scope of their works e.g., [@millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring]. Indeed, while accessibility can be seen as the *potential* for spatial interaction--and Wilson [@wilson1971] briefly touches on this--such mentions have not resulted in deeper analytical integration of these concepts. Furthermore, some of this literature also occasionally conflates or blurs the distinction entirely, for instance by co-citing Hansen and Wilson as being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending]. This conflation reveals ongoing murkiness between the distinction of spatial interaction and the *potential for* spatial interaction in the literature.

@@

On that note, Soukhov et al., 2023 and 2024 [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024] introduce the balancing factors as a mechanisms to ensure that opportunities at each destination are proportionally allocated to each zone (based on the proportion of population seeking opportunities and the relative travel impedance). This is to ensure that each zonal accessibility value is the sum of this proportional allocation from each destination, and that all zonal values ultimately sum to the total number of opportunities in the region. However, these balancing factors were deduced intuitively. These works do not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly constrained model (derived from entropy maximization). This equivalence is only discovered in hindsight, as will be demonstrated in the following section. These two works also do not discuss other constrained cases that will also be addressed.

In sum, despite the interpretative advantaged offered by the statistical logic of Wilson [@wilson1971]'s framework, the spatial interaction-oriented Hansen [@hansen1959] citing literature nor the accessibility-oriented literature that cite Wilson [@wilson1971] have yet to meaningfully adopt Wilson's constraint-based logic to the concept of accessibility. So, in the next section, we demonstrate how applying Wilson’s framework enables accessibility to move from a proportional relationship to calibrated equality--tying outputs to tangible system knowns. This re-expression of accessibility using constraints yields interpretable, unit-consistent measures of opportunity. This approach takes the same path of entropy-maximisation as in Wilson [@wilson1971], and does not rely on specifying some universal constant $G$ like initially proposed in Stewart [@stewartDemographicGravitationEvidence1948] (recall, @eq-stewart-population-potential-sum which Hansen [@hansen1959] operationalised).

@@

Despite the close conceptual ties between accessibility and spatial interaction modeling, the former has not meaningfully absorbed the constraint-based logic of the latter. This section introduces this reframing: defining a family of accessibility measures using Wilson’s approach grounded in statistical mechanics, shifting place-based accessibility (a la Hansen [@hansen1959]) from a relationship describing proportionality to equality.

This shift addresses the fundamental issue of interpretability associated with the proportional nature of Hansen-type accessibility indicators we previous outlined. Rather than relying on arbitrary or uncalibrated proportional relationships (or invoking a universal constant $G$ as proposed in Stewart [@stewartDemographicGravitationEvidence1948]) we adopt Wilson’s approach grounded in statistical mechanics.

>>>>

Firstly, there is a group of articles within this stream that cite Wilson [@wilson1971] exclusively as attribution for using context-dependent travel cost functions. This trend is common: for instance it is done in [@handyMeasuringAccessibilityExploration1997; @weibullNumericalMeasurementAccessibility1980; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining]. However, these works do not engage with spatial interaction beyond this attribution.

Secondly, a subset of literature explicitly associates spatial interaction, as defined in Wilson [@wilson1971], with accessibility’s potential for spatial interaction--but only superficially. These works acknowledge the conceptual link, but do not go beyond this recognition in the scope of their works e.g., [@grengs2010job; @millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring]. Indeed, while accessibility can be seen as the *potential* for spatial interaction--and Wilson [@wilson1971] briefly touches on this--such mentions have not resulted in deeper analytical integration of these concepts. Furthermore, some of this literature also occasionally conflates or blurs the distinction entirely, for instance by co-citing Hansen and Wilson as being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending]. This conflation reveals ongoing murkiness between the distinction of spatial interaction and the *potential for* spatial interaction in the literature.

@@

On that note, Soukhov et al., 2023 and 2024 [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024] introduce the balancing factors as a mechanisms to ensure that opportunities at each destination are proportionally allocated to each zone (based on the proportion of population seeking opportunities and the relative travel impedance). This is to ensure that each zonal accessibility value is the sum of this proportional allocation from each destination, and that all zonal values ultimately sum to the total number of opportunities in the region. However, these balancing factors were deduced intuitively. These works did not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly constrained model (derived from entropy maximization). This equivalence is only discovered in hindsight, as will be demonstrated in the following section. These two works also do not discuss other constrained cases that will also be addressed.

In sum, despite the interpretative advantaged offered by the statistical logic of Wilson [@wilson1971]'s framework, neither the spatial interaction literature that cites Hansen [@hansen1959] nor the accessibility literature citing Wilson [@wilson1971] has meaningfully applied Wilson's constraint-based logic to the concept of accessibility. So, in the next section, we demonstrate how applying Wilson’s framework enables accessibility to move from a proportional relationship to calibrated equality--tying outputs to tangible system knowns. This re-expression of accessibility using constraints yields interpretable, unit-consistent measures of opportunity. This approach takes the same path of entropy-maximisation as in Wilson [@wilson1971], and does not rely on specifying some universal constant $G$ like initially proposed in Stewart [@stewartDemographicGravitationEvidence1948] (recall, @eq-stewart-population-potential-sum which Hansen [@hansen1959] operationalised).

@@

Despite the close conceptual ties between accessibility and spatial interaction modeling, the former has not meaningfully absorbed the constraint-based logic of the latter. This section introduces this reframing: defining a family of accessibility measures using Wilson’s approach grounded in statistical mechanics, shifting place-based accessibility (a la Hansen [@hansen1959]) from a relationship describing proportionality to equality. This shift addresses the fundamental issue of interpretability associated with the proportional nature of Hansen-type accessibility indicators we previous outlined.
